---
title: "ARRA Analysis: NYC"
output:
  html_document: default
---

American Recovery and Reinvestment Act of 2009 was passed on 2009-02-17.

> "Making supplemental appropriations for job preservation and creation, infrastructure investment, energy efficiency and science, assistance to the unemployed, and State and local fiscal stabilization, for the fiscal year ending September 30, 2009, and for other purposes.

> (a) Statement of Purposes.--The purposes of this Act include the following:
>
> (1) To preserve and create jobs and promote economic recovery.
>
> (2) To assist those most impacted by the recession.
>
> (3) To provide investments needed to increase economic efficiency by spurring technological advances in science and health.
>
> (4) To invest in transportation, environmental protection, and other infrastructure that will provide long-term economic benefits.
>
> (5) To stabilize State and local government budgets, in order to minimize and avoid reductions in essential services and counterproductive state and local tax increases.


## Potential Questions


Low hanging fruit

- recreate the NYC Mayor's Office of Operation's midway report with the final data. [Link is a direct download to a pdf](http://www.nyc.gov/html/ops/nycstim/downloads/pdf/human_services_council_presentation_mayors_office_of_operations.pdf) (slides 17, and 22-24 have the relevant figures).

Higher up fruit

- Was the ARRA funding supplemented with city funding for larger projects?
- What was the distribution of funding by sector? and what were the biggest projects in each sector? 
- Were projects completed on time anmd in budget?
- Mine Payment Description (text field) by sector to see what the money was being used on? 


The NYC Mayor's Office of Operations used four categories

Recovery (job creation)

Relief (assist those most struggling, stabilize state and local budgets)

- Where was the city budget supplemented, where was it displaced?
- What was the job market like in NYC before and after the ARRA funding (unemployment, number of jobs, type of jobs)?

Reinvestment (infrastructure, education, and environment)

- What was built? What were the investments in education? 

Responsibility (Regular reporting, transparency)

- Check payment dates, contract dates, large lump sum payments



```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE) 
```



```{r}
# Load Packages
# First we will install a package manager: pacman
if (!require("pacman")) {install.packages("pacman")} 

# And then we will use pacman to check if the other needed packages are installed and install if necessary

# Data manipulation
pacman::p_load(tidyverse) # General R improvements includes stringr, dplyr
pacman::p_load(lubridate) # date manipulation
pacman::p_load(stringi) # string processing, has some functions that work better than stringr
# pacman::p_load(data.table)   # extension of data.frame, has in-place replacement
pacman::p_load(countrycode) # converts country names to country codes, used in affiliation reconciliation

# Data access
pacman::p_load(filesstrings) # make and edit files/folders
pacman::p_load(httr) # api access
pacman::p_load(xml2) # xml access and parsing
pacman::p_load(jsonlite) # JSON interpreter
pacman::p_load(readxl) # read in excel files

# Data visualization
pacman::p_load(ggplot2) # plotting package, not used currently but can be used to look at the networks visually
pacman::p_load(scales) # plotting package, extends ggplot2
# pacman::p_load(RColorBrewer) # extra color palettes

# Misc
pacman::p_load(igraph) # networks: creation, transformation, etc

# Restart R session if you install new packages
# .rs.restartR()


```

Next we have to point to where the data lives in the repository. 
It is in a ./Data/ folder but in not in the git repo. 
The data can be found on [kaggle](https://www.kaggle.com/datasets/new-york-city/nyc-use-of-arra-stimulus-funds?resource=download) or on NYC's city [OpenData webpage](https://data.cityofnewyork.us/City-Government/Use-of-ARRA-Stimulus-Funds/ivix-m77e)

The dataset covers the use of funds from the American Recovery and Reinvestment Act of 2009.
The dataset was originally made public in 2011.
It was last updated in 2018 and the metadata was last updated in 2022.

``` {r}
dataFolder <- "./Data/"

ArraDataFile <- "use-of-arra-stimulus-funds.csv" 
ArraJSONFile <- "socrata_metadata.json"  
ArradDictionaryFile <- "Data_Dictionary_Use_of_ARRA_Stimulus_Funds.xlsx"


```

## Load data

Read in data and the metadata.

``` {r}
# Read in the data file
arraData <- read.csv(file = paste0(dataFolder, ArraDataFile)) %>% as_tibble
# I'm choosing to import and convert to tibble to make use of dplyr's filtering and ease of use with ggplot2
# I could also have used data.table which has better in place replacement but since this dataset is not huge and I do not expect to need in place replacement, tibble's are a better fit.


# Read in the metadata file
arraMetaData <- read_json(path = paste0(dataFolder, ArraJSONFile), simplifyVector = FALSE) 

```



Convert date columns from strings to dates.

``` {r}
arraData$Actual.Completion.Date <- arraData$Actual.Completion.Date %>% as_date()
arraData$Date.Funds.Awarded.by.Fed.State <- arraData$Date.Funds.Awarded.by.Fed.State %>% as.Date()
arraData$Actual.Start.Date <- arraData$Actual.Start.Date %>% as.Date()
arraData$Payment.Date <- arraData$Payment.Date %>% as.Date()

```



Some initial thoughts
There are `r arraData$Stimulus.Tracker.ID %>% unique() %>% length()` unique tracked projects in `r arraData$Funding.Category %>% unique() %>% length()` categories.
Awards were awarded through `r length(unique(arraData$Award.Lead.City.Agency))` different city agencies. 


There are `r arraData$Payment.Id %>% length()` rows.
The unique id column in each row is the Payment ID with `r (arraData$Payment.Id %>% unique() %>% length())-1` unique entries and `r which(is.na(arraData$Payment.Id))%>% length()` with no value.
The NAs are probably worth looking into at some point. 


# Pie charts

Recreate one of the pie charts from the midterm report.
This is pie chart covers the distribution of the funds over the eight sectors


``` {r}

stimIdSum <- arraData %>% 
  filter(is.na(Payment.Id)==FALSE) %>%
  group_by(Stimulus.Tracker.ID,Funding.Category) %>% 
  summarise(TotalMoney = sum(Payment.Value), .groups= "keep") %>%
  arrange(desc(TotalMoney) ) 

# head(stimIdSum)

```



``` {r}

stimIdSum %>% 
  group_by(Funding.Category) %>%
  summarize(Money = sum(TotalMoney)) %>%
  ggplot( aes(x="", y=Money, fill=Funding.Category ))+
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0) +
  ggtitle("Proportion money paid out by category")+
  theme_void() # remove background, grid, numeric labels


```


Which we can compare to the number of payments made to get a sense of whether the payments were in large lump sums or more spread out. 

``` {r}
arraData %>% 
  filter(is.na(Payment.Id)==FALSE) %>%
  count(Funding.Category)  %>%
  ggplot( aes(x="", y=n, fill=Funding.Category ))+
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0) +
  ggtitle("Proportion of the number of payments by category")+
  theme_void() # remove background, grid, numeric labels


```


The order of the funding categories is the same in the two pie charts.
Medicaid relief while being the largest category was done is 8 payments.  



# Total money awarded

The total of the payment value column is `r round(sum(arraData$Payment.Value, na.rm = TRUE)/10^6)` million dollars.
The number listed in the presentation referenced above my the Mayors office lists that as of  December 31, 2009 they have: \$6,034 million awarded, received \$185M, and spent \$475M.




# Funding size


``` {r}
arraCompleted <- arraData %>% filter(Project.Status == "Fully Completed"  )

LargePayments <- arraCompleted %>% 
  filter(Payment.Value > 1000000) %>%
  arrange(desc(Payment.Value))

# List the individual payments over 1 million dollars. 
print(LargePayments[,c("Payment.Value","Project.Name", "Funding.Category")])

```


Only `r arraCompleted$Stimulus.Tracker.ID %>% unique %>% length()` projects are listed as "Fully Completed."
`r LargePayments$Project.Name %>% length()` of them had final payments over 1 million dollars and those payments come from `r LargePayments$Stimulus.Tracker.ID %>% unique() %>% length()` different projects.







``` {r}
# Add a color blind friendly palette
# ref: http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/#a-colorblind-friendly-palette
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73",  "#0072B2","#F0E442", "#D55E00", "#CC79A7")


stimIdSum  %>% 
  mutate(TotalMoney = ifelse(TotalMoney > 2000000, 2000000, TotalMoney)) %>% # If the total money for a project is great than 2 million set it to 2 million for the sake of seeing it on the histogram
  ggplot() +
  geom_histogram(aes(x = (TotalMoney/1000), fill = Funding.Category),
                 breaks=seq(0, 2000 , by = 50),
                 # colour="black",
                 # fill="purple"
  )+
  scale_fill_manual(values=cbPalette)+
  theme(
    plot.background = element_rect(fill = "#D9D9D9")
    , legend.key = element_rect(fill = "#D9D9D9")
    , legend.background = element_rect(fill="#D9D9D9")
    
    , panel.background = element_rect(fill = "#D9D9D9",
                                      colour = "#D9D9D9",
                                      size = 0.5, linetype = "solid")
    , panel.border = element_rect(fill = "NA",
                                  colour = "black",
                                  size = 0.5, linetype = "solid")
    , panel.grid.major = element_blank()
    , panel.grid.minor = element_blank()
    , axis.ticks.length = -unit(0, "pt")
    # , axis.title.y = element_text(angle = 0, vjust = 0.5, hjust=0)
  )+
  scale_x_continuous(limits=c(0, 2100), expand = c(0, 0)) +
  scale_y_continuous(limits=c(0, 50), expand = c(0, 0)) +
  theme(plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"))+
  xlab("Ammount per project ($ in Thousands)")+
  ylab("Number of Projects")



ggplot(stimIdSum) +
  geom_histogram(aes(x = (TotalMoney/1000), fill = Funding.Category),
                 breaks=seq(0, 1500 , by = 50),
                 # colour="black",
                 # fill="purple"
  )+
  scale_fill_manual(values=cbPalette)+
  theme(
    plot.background = element_rect(fill = "#D9D9D9")
    , legend.key = element_rect(fill = "#D9D9D9")
    , legend.background = element_rect(fill="#D9D9D9")
    
    , panel.background = element_rect(fill = "#D9D9D9",
                                      colour = "#D9D9D9",
                                      size = 0.5, linetype = "solid")
    , panel.border = element_rect(fill = "NA",
                                  colour = "black",
                                  size = 0.5, linetype = "solid")
    , panel.grid.major = element_blank()
    , panel.grid.minor = element_blank()
    , axis.ticks.length = -unit(0, "pt")
    # , axis.title.y = element_text(angle = 0, vjust = 0.5, hjust=0)
  )+
  scale_x_continuous(limits=c(0, 1500), expand = c(0, 0)) +
  scale_y_continuous(limits=c(0, 15), expand = c(0, 0)) +
  theme(plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"))+
  xlab("Ammount per project ($ in Thousands)")+
  ylab("Number of Projects")


```


# Time of Payments

Look into the timing of the payments. 
Since there were fairly strict reporting requirements for ARRA, ideally there should be no obvious gaps. 


First lets look at the completion date for each project. 

``` {r}
completionCounts <- count(arraData, Actual.Completion.Date)%>%
  arrange(desc(n) ) 

completionCounts
# monthdf <- tibble(Month = month(arraData$Actual.Completion.Date), 
#                   Year = year(arraData$Actual.Completion.Date))
# 
# ggplot(monthdf) +
#   geom_histogram(aes(x = Month),
#                  breaks=seq(1, 12, by = 1),
#                  colour="black",
#                  fill="purple"
#   )
# 
# ggplot(datedf) +
#   geom_histogram(aes(x = Year),
#                  breaks=seq(2008, 2022 , by = 1),
#                  colour="black",
#                  fill="purple"
#   )


```


There are only `r length(completionCounts$n)` days on which projects were listed as completed.


``` {r}
paymentDates <- arraData %>% 
  filter(is.na(Payment.Id)==FALSE) %>%
  group_by(Payment.Date, Funding.Category) %>% 
  summarise(TotalMoney = sum(Payment.Value), .groups= "keep") 




ggplot(paymentDates) +
  geom_point(aes(Payment.Date, (TotalMoney/1000), colour = Funding.Category)
  )+
  # scale_color_manual(values=cbPalette)+
  theme(
    plot.background = element_rect(fill = "#D9D9D9")
    , legend.key = element_rect(fill = "#D9D9D9")
    , legend.background = element_rect(fill="#D9D9D9")
    
    , panel.background = element_rect(fill = "#D9D9D9",
                                      colour = "#D9D9D9",
                                      size = 0.5, linetype = "solid")
    , panel.border = element_rect(fill = "NA",
                                  colour = "black",
                                  size = 0.5, linetype = "solid")
    , panel.grid.major = element_blank()
    , panel.grid.minor = element_blank()
    , axis.ticks.length = -unit(0, "pt")
    # , axis.title.y = element_text(angle = 0, vjust = 0.5, hjust=0)
  )+
  theme(plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"))+
  xlab("Date")+
  ylab("Amount spent per day ($ in Thousands)")



```




``` {r}

ggplot(paymentDates) +
  geom_point(aes(Payment.Date, TotalMoney, colour = Funding.Category)
  )+
  scale_y_continuous(trans='log10'
                     , breaks=breaks_log()
                     , labels=trans_format('log10', math_format(10^.x))
                     , limits = c(10^0, 10^9)
                     , expand = c(0, 0)
                     , sec.axis = dup_axis(trans = ~.,
                                           name = NULL,
                                           breaks = derive(),
                                           labels = NULL,
                                           guide = derive()
                     )
  )+
  theme(
    plot.background = element_rect(fill = "#D9D9D9")
    , legend.key = element_rect(fill = "#D9D9D9")
    , legend.background = element_rect(fill="#D9D9D9")
    
    , panel.background = element_rect(fill = "#D9D9D9",
                                      colour = "#D9D9D9",
                                      size = 0.5, linetype = "solid")
    , panel.border = element_rect(fill = "NA",
                                  colour = "black",
                                  size = 0.5, linetype = "solid")
    , panel.grid.major = element_blank()
    , panel.grid.minor = element_blank()
    , axis.ticks.length = -unit(0, "pt")
    # , axis.title.y = element_text(angle = 0, vjust = 0.5, hjust=0)
  )+
  theme(plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"))+
  xlab("Date")+
  ylab("Amount spent per day ($)")


```



``` {r}

paymentDatesWide <- arraData %>% 
  filter(is.na(Payment.Id)==FALSE) %>%
  group_by(Payment.Date, Funding.Category) %>% 
  summarise(TotalMoney = sum(Payment.Value), .groups= "keep") %>%
  pivot_wider(names_from = Funding.Category, values_from = TotalMoney)



ggplot(paymentDatesWide) +
  geom_point(aes(x = Payment.Date, y = (`Infrastructure`/1000 ))
  )+
  theme(
    plot.background = element_rect(fill = "#D9D9D9")
    
    , panel.background = element_rect(fill = "#D9D9D9",
                                      colour = "#D9D9D9",
                                      size = 0.5, linetype = "solid")
    , panel.border = element_rect(fill = "NA",
                                  colour = "black",
                                  size = 0.5, linetype = "solid")
    , panel.grid.major = element_blank()
    , panel.grid.minor = element_blank()
    , axis.ticks.length = -unit(0, "pt")
    , axis.title.y = element_text(angle = 0, vjust = 0.5, hjust=0)
  )+
  # scale_x_continuous(limits=c(0, 1500), expand = c(0, 0)) +
  # scale_y_continuous(limits=c(0, 15), expand = c(0, 0)) +
  theme(plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"))+
  xlab("Date")+
  ylab("Total money spent by day: Infrastructure")



```


# Network analysis

``` {r}
tmp <-  arraData %>% 
  filter(is.na(Payment.Id)==FALSE)
# tmp$Funding.Category
# tmp$Award.Lead.City.Agency
# 
# tmp$Project.Lead.City.Agency

network <- tmp[,c("Funding.Category", "Project.Lead.City.Agency")] %>%
  # group_by(Funding.Category, Project.Lead.City.Agency) %>% 
  graph.data.frame( directed=FALSE) %>%
  simplify( edge.attr.comb=list( "ignore")) # Combines duplicate edges

# vertex_attr_names(network)
# edge_attr_names(network)


# coMatrix <- as_adjacency_matrix(network)

V(network)$level <- 0
categories <- arraData$Funding.Category %>% unique()

for( netiter in 1:length(V(network)$name)){
  if (V(network)[netiter]$name %in% categories ){
    V(network)[netiter]$level <- 1
  }
  
}

# Sugiyama bipartite network layout
coords = layout_with_sugiyama(network, layers = V(network)$level, hgap = 10, vgap = 1,
                              maxiter = 100, weights = NULL, attributes = c("default", "all", "none"))


par(bg = "#D9D9D9")
plot(network
     , layout =-coords$layout[,2:1]
     , vertex.color = ifelse(V(network)$level ==  1, "#0F3F66", "yellow"  )
     , vertex.label.degree = pi
     , vertex.label.dist=ifelse(V(network)$level ==  1, -10, 10 )
     #, vertex.label=  NA  
     , vertex.size = 4
     #, margin =c(0,0,0,0)
     #, asp = 0
)
```

The edges have been combined so as to see the structure. 
This does throw out interesting information like the strength of the connections (amount of funding, number of payments or projects).
This could also have been represented as a matrix with the departments along one side and the categories along the other. 



``` {r}

coords = layout_with_fr(network) # Force directed network layout 

par(bg = "#D9D9D9")
plot(network
     , layout =coords
     , vertex.color = ifelse(V(network)$level ==  1, "#0F3F66", "yellow"  )
     , vertex.label.degree = pi
     , vertex.label.dist=ifelse(V(network)$level ==  1, -10, 10 )
     , vertex.label=  NA  
     , vertex.size = 4
     #, margin =c(0,0,0,0),
     # , asp = 0
)

```


Blue nodes are the categories and yellow are the city departments.
The labels are not shown so the structure can be seen.
Only a few departments work in multiple categories.

Visually there are some obvious clusters and one next step could be enumerating them. 
Another next step could be to repeat this with Award Lead City Agency instead of Project Lead City Agency in case they are different.
Yet another would be to look at the strength of the connections based on number of payments, number of projects, or amount of funding. 

